---
title: "2_primary_anlaysis"
author: "Joy_Fu"
date: "2023-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
rm(list = ls())
lapply(paste('package:', names(sessionInfo()$otherPkgs), sep = ""),
       detach, character.only = TRUE, unload = TRUE)
pacman::p_load(tidyverse, gtsummary, epiflow, gtools, caret, gridExtra, forestplot, h2o)
raw_data_path = "/Users/Mingzhou/Desktop/Projects/AD-LOE-genetics/data/"
output_path = "/Users/Mingzhou/Desktop/Projects/AD-LOE-genetics/output/"
# Source in useful functions
source("functions.R")

# Load in datasets
load(file = paste0(raw_data_path, "modeling/EUR/patient_final_EUR.rda"))
load(file = paste0(raw_data_path, "modeling/EUR/freeze60k_geno_freq.rda"))
load(file = paste0(raw_data_path, "modeling/EUR/genotype_df_map.rda"))
```

# Part 1. Finalize analytical sample
```{r}
# only include patients with valid AD and epilepsy conditions
final_sample_EUR = patient_final_EUR %>% 
  mutate(htn = if_else(!is.na(HTNDate), 1, 0),
         diabetes = if_else(!is.na(DiabetesDate), 1, 0),
         stroke = if_else(!is.na(StrokeDate), 1, 0),
         hyperlipid = if_else(!is.na(HyperlipidDate), 1, 0)) %>% 
  mutate(either = case_when(
    AD == 0 & epilepsy == 0 ~ 0,
    TRUE ~ 1
  )) %>% 
  mutate(both = case_when(
    AD == 1 & epilepsy == 1 ~ 1,
    AD == 0 & epilepsy == 0 ~ 0
  )) %>% 
  filter(!is.na(either)) %>% filter(!is.na(AD) | !is.na(epilepsy)) %>% 
  select(UniqueSampleId, AD, epilepsy, ad_age_diagnosis, epi_age_diagnosis, 
         age_last_visit, female, PC1, PC2, PC3, PC4, record_length, enc_per_yr, 
         htn, diabetes, stroke, hyperlipid, dementia, e4count,
         AD_PRS_full_norm, AD_PRS_rmapoe_norm) 
dim(final_sample_EUR) # dim = (17031,23)
```

# Part 2. Descriptive statistics
```{r}
# Desc by epilepsy status
final_sample_EUR %>% mutate(epi.cat = as.factor(epilepsy)) %>% 
  select(epi.cat, AD, either, age_last_visit, female, record_length, enc_per_yr, 
         htn, diabetes, stroke, hyperlipid, dementia, e4count, AD_PRS_full_norm, 
         AD_PRS_rmapoe_norm) %>% 
  tbl_summary(by = epi.cat) %>% add_p()
```

```{r}
# Desc by ad status
final_sample_EUR %>% mutate(AD.cat = as.factor(AD)) %>% 
  select(AD.cat, epilepsy, either, age_last_visit, female, record_length, enc_per_yr, 
         htn, diabetes, stroke, hyperlipid, dementia, e4count, AD_PRS_full_norm, 
         AD_PRS_rmapoe_norm) %>% 
  tbl_summary(by = AD.cat) %>% add_p()
```


# Part 3. Association tests
## 1. APOE (e4count)
### LOE
```{r}
demo_basic = 'age_last_visit + female + PC1 + PC2 + PC3 + PC4 + record_length + enc_per_yr'
health_info = paste0(demo_basic, ' + htn + diabetes + hyperlipid') 

# Demographic only
demo_only = glm(epilepsy ~ age_last_visit + female + record_length + enc_per_yr, 
                data = final_sample_EUR, family = 'binomial')
# Adjusted for basic demo
e4count_demo = glm(as.formula(paste('epilepsy ~ e4count + ', demo_basic)), 
                   data = final_sample_EUR, family = 'binomial')
# Adjusted for health conditions
e4count_health = glm(as.formula(paste('epilepsy ~ e4count + ', health_info)), 
                     data = final_sample_EUR, family = 'binomial')
```

### AD
```{r}
# Demographic only
demo_only = glm(AD ~ age_last_visit + female + record_length + enc_per_yr, 
                data = final_sample_EUR, family = 'binomial')
# Adjusted for basic demo
e4count_demo = glm(as.formula(paste('AD ~ e4count + ', demo_basic)), 
                   data = final_sample_EUR, family = 'binomial')
# Adjusted for health conditions
e4count_health = glm(as.formula(paste('AD ~ e4count + ', health_info)), 
                     data = final_sample_EUR, family = 'binomial')
```


## 2. AD PRS full
### LOE
```{r}
# Adjusted for basic demo
prs_full_demo = glm(as.formula(paste('epilepsy ~ AD_PRS_full_norm + ', demo_basic)), 
                    data = final_sample_EUR, family = 'binomial')
# Adjusted for health conditions
prs_full_health = glm(as.formula(paste('epilepsy ~ AD_PRS_full_norm + ', health_info)), 
                      data = final_sample_EUR, family = 'binomial')
```

### AD
```{r}
# Adjusted for basic demo
prs_full_demo = glm(as.formula(paste('AD ~ AD_PRS_full_norm + ', demo_basic)), 
                    data = final_sample_EUR, family = 'binomial')
# Adjusted for health conditions
prs_full_health = glm(as.formula(paste('AD ~ AD_PRS_full_norm + ', health_info)), 
                      data = final_sample_EUR, family = 'binomial')
```

## 3. AD PRS w/o APOE
### LOE
```{r}
# Adjusted for basic demo
prs_rmapoe_demo = glm(as.formula(paste('epilepsy ~ AD_PRS_rmapoe_norm + ', demo_basic)), 
                      data = final_sample_EUR, family = 'binomial')
# Adjusted for health conditions
prs_rmapoe_health = glm(as.formula(paste('epilepsy ~ AD_PRS_rmapoe_norm + ', health_info)), 
                        data = final_sample_EUR, family = 'binomial')
```

### AD
```{r}
# Adjusted for basic demo
prs_rmapoe_demo = glm(as.formula(paste('AD ~ AD_PRS_rmapoe_norm + ', demo_basic)), 
                      data = final_sample_EUR, family = 'binomial')
# Adjusted for health conditions
prs_rmapoe_health = glm(as.formula(paste('AD ~ AD_PRS_rmapoe_norm + ', health_info)), 
                        data = final_sample_EUR, family = 'binomial')
```

## 4. AD and LOE
### LOE
```{r}
# Adjusted for basic demo
loe_ad_demo = glm(as.formula(paste('epilepsy ~ AD + ', demo_basic)), 
                  data = final_sample_EUR, family = 'binomial')
# Adjusted for health conditions
loe_ad_health = glm(as.formula(paste('epilepsy ~ AD + ', health_info)), 
                    data = final_sample_EUR, family = 'binomial')
```

### AD
```{r}
# Adjusted for basic demo
ad_loe_demo = glm(as.formula(paste('AD ~ epilepsy + ', demo_basic)), 
                  data = final_sample_EUR, family = 'binomial')
# Adjusted for health conditions
ad_loe_health = glm(as.formula(paste('AD ~ epilepsy + ', health_info)), 
                    data = final_sample_EUR, family = 'binomial')
```


```{r message=FALSE, warning=FALSE}
# Run all models to get results
lst_model = c('demo_only', 'e4count_demo', 'e4count_health', 'prs_full_demo',
              'prs_full_health', 'prs_rmapoe_demo', 'prs_rmapoe_health',
              'loe_ad_demo', 'loe_ad_health')
loe_prs = make_OR_table(lst_model, 2, 'loe_')

lst_model = c('demo_only', 'e4count_demo', 'e4count_health', 'prs_full_demo',
              'prs_full_health', 'prs_rmapoe_demo', 'prs_rmapoe_health',
              'ad_loe_demo', 'ad_loe_health')
ad_prs = make_OR_table(lst_model, 2, 'ad_')
```


# Part 4. Important risk SNPs modeling
```{r}
demo_set = c("female", "age_last_visit", "record_length", "enc_per_yr", 
             "PC1", "PC2", "PC3", "PC4")
health_set = c(demo_set, "htn", "diabetes", "hyperlipid")
load(file = paste0(raw_data_path, "modeling/EUR/mapped_snp.rda"))
mapped_AD_snp = mapped_snp %>% filter(rank_CADD == 1) %>% 
  filter(phenotype == "AD") %>% unique()
model_df_prep = freeze60k_geno_freq %>%
  rownames_to_column(var = "UniqueSampleId") %>% 
  mutate(UniqueSampleId = as.numeric(UniqueSampleId)) %>% 
  inner_join(final_sample_EUR) %>% 
  select(UniqueSampleId, all_of(names(genotype_df_map)), all_of(health_set), 
         epilepsy, AD, either, AD_PRS_full_norm, AD_PRS_rmapoe_norm, e4count) %>% 
  column_to_rownames(var = "UniqueSampleId")
dim(model_df_prep) # dim = (17031,179)
h2o.init(nthreads = -1)
seed_num = 4766
```

## 1. Predict AD | LOE
```{r}
Y = "either"
target.train = as.h2o(model_df_prep)
target.train[,Y] = as.factor(target.train[,Y])
X = c(demo_set, mapped_AD_snp$uniqueID)
# Step 1: feature selection
SNP.lasso.either = h2o.glm(training_frame = target.train, x = X, y = Y, 
                           balance_classes = T, model_id = "LASSO either", 
                           nfolds = 5, fold_assignment = "Stratified", 
                           seed = seed_num, family = "binomial", 
                           alpha = 1, lambda_search = TRUE, standardize = T, 
                           keep_cross_validation_models = F)
snps_either = SNP.lasso.either@model$variable_importances %>% 
  as.data.frame() %>% filter(variable %!in% health_set) %>% 
  filter(scaled_importance > 0) %>% pull(variable)
X = c(demo_set, snps_either)
SNP.lasso.either.pred = h2o.glm(training_frame = target.train, x = X, y = Y, 
                                balance_classes = T, model_id = "LASSO either pred", 
                                nfolds = 5, fold_assignment = "Stratified", 
                                seed = seed_num, family = "binomial", lambda = 0, 
                                standardize = T, keep_cross_validation_models = F)
either.pred = h2o.predict(SNP.lasso.either.pred, target.train) %>% as.data.frame()
```

```{r}
new_model_data = cbind(model_df_prep, either.pred$p1) 
names(new_model_data)[180] = "pred_score"
```

## 2. Predict AD
```{r}
# Set response variable
Y = "AD"
# set h2o project
target.train = as.h2o(new_model_data)
target.train[,Y] = as.factor(target.train[,Y])
```

### 1) Comparison model
```{r}
#========= Benchmark 1: demo only =========
X = c("age_last_visit", "female", "record_length", "enc_per_yr")
demo.only = h2o.glm(training_frame = target.train, x = X, y = Y, 
                    balance_classes = T, model_id = "Demo only", 
                    nfolds = 5, fold_assignment = "Stratified",
                    seed = seed_num, family = "binomial", lambda = 0, 
                    standardize = T, keep_cross_validation_models = F)
#========= AD PRS model =========
X = c("AD_PRS_full_norm", demo_set)
PRS.full = h2o.glm(training_frame = target.train, x = X, y = Y, 
                   balance_classes = T, model_id = "AD PRS full", 
                   nfolds = 5, fold_assignment = "Stratified", 
                   seed = seed_num, family = "binomial", lambda = 0, 
                   standardize = T, keep_cross_validation_models = F)
X = c("AD_PRS_rmapoe_norm", demo_set)
PRS.rmapoe = h2o.glm(training_frame = target.train, x = X, y = Y, 
                     balance_classes = T, model_id = "AD PRS without APOE", 
                     nfolds = 5, fold_assignment = "Stratified", 
                     seed = seed_num, family = "binomial", lambda = 0, 
                     standardize = T, keep_cross_validation_models = F)
X = c("e4count", demo_set)
e4count = h2o.glm(training_frame = target.train, x = X, y = Y, 
                  balance_classes = T, model_id = "e4count", 
                  nfolds = 5, fold_assignment = "Stratified", 
                  seed = seed_num, family = "binomial", lambda = 0, 
                  standardize = T, keep_cross_validation_models = F)
```

### 2) Direct LASSO
```{r}
X = c(mapped_AD_snp$uniqueID, demo_set)
snp.lasso.AD.feature = h2o.glm(training_frame = target.train, x = X, y = Y, 
                               balance_classes = T, model_id = "LASSO AD feature", 
                               nfolds = 5, fold_assignment = "Stratified", 
                               seed = seed_num, family = "binomial",
                               alpha = 1, lambda_search = TRUE,
                               standardize = T, keep_cross_validation_models = F)
snps_AD = snp.lasso.AD.feature@model$variable_importances %>% 
  as.data.frame() %>% filter(variable %!in% demo_set) %>% 
  filter(scaled_importance > 0) %>% pull(variable)
X = c(snps_AD, demo_set)
snp.direct.lasso.AD = h2o.glm(training_frame = target.train, x = X, y = Y, 
                              balance_classes = T, model_id = "LASSO direct AD", 
                              nfolds = 5, fold_assignment = "Stratified", 
                              seed = seed_num, family = "binomial", lambda = 0, 
                              standardize = T, keep_cross_validation_models = F)
```

### 3) Two-step LASSO
```{r}
X = c(mapped_AD_snp$uniqueID, demo_set, "pred_score")
snp.lasso2.AD.feature = h2o.glm(training_frame = target.train, x = X, y = Y, 
                                balance_classes = T, model_id = "LASSO AD feature 2", 
                                nfolds = 5, fold_assignment = "Stratified", 
                                seed = seed_num, family = "binomial",
                                alpha = 1, lambda_search = TRUE,
                                standardize = T, keep_cross_validation_models = F)
snps_AD_specific = snp.lasso2.AD.feature@model$variable_importances %>% 
  as.data.frame() %>% filter(variable %!in% demo_set) %>% 
  filter(scaled_importance > 0) %>% pull(variable)
X = c(snps_AD_specific, demo_set)
snp.twostep.lasso.AD = h2o.glm(training_frame = target.train, x = X, y = Y, 
                               balance_classes = T, model_id = "LASSO two-step AD", 
                               nfolds = 5, fold_assignment = "Stratified", 
                               seed = seed_num, family = "binomial", lambda = 0, 
                               standardize = T, keep_cross_validation_models = F)
```

```{r}
all_model_metric = c()
all_model_performance = c()
# Extract results
model_list = c("demo.only", "PRS.full", "PRS.rmapoe", "e4count",
               "snp.direct.lasso.AD", "snp.twostep.lasso.AD")
for (i in 1:length(model_list)) {
  print(paste0("Model: ", model_list[i]))
  # Run extraction function
  extract_cv = extract_cv_results(get(model_list[i]))
  model_metric = t(extract_cv[[1]]) %>% as.data.frame() %>% 
    mutate(mark = model_list[i])
  model_performance = extract_cv[[2]] %>% as.data.frame() %>% 
    mutate(mark = model_list[i])
  if (i == 1) {
    all_model_metric = model_metric
    all_model_performance = model_performance
  } else {
    all_model_metric = rbind(all_model_metric, model_metric)
    all_model_performance = rbind(all_model_performance, model_performance)
  }
}
# Join results together
model_metric_AD = rbind(all_model_metric)
model_performance_AD = rbind(all_model_performance)
model_summary_AD = model_metric_AD %>% inner_join(model_performance_AD)
write.table(model_summary_AD,
            file = paste0(output_path, "model_summary_AD.txt"),
            sep = "\t", quote = F, row.names = F, col.names = T)
```

## 3. Predict epilepsy
```{r}
# Set response variable
Y = "epilepsy"
# set h2o project
target.train = as.h2o(new_model_data)
target.train[,Y] = as.factor(target.train[,Y])
```

### 1) Comparison model
```{r}
#========= Benchmark 1: demo only =========
X = c("age_last_visit", "female", "record_length", "enc_per_yr")
demo.only = h2o.glm(training_frame = target.train, x = X, y = Y, 
                    balance_classes = T, model_id = "Demo only", 
                    nfolds = 5, fold_assignment = "Stratified",
                    seed = seed_num, family = "binomial", lambda = 0, 
                    standardize = T, keep_cross_validation_models = F)
#========= AD PRS model =========
X = c("AD_PRS_full_norm", demo_set)
PRS.full = h2o.glm(training_frame = target.train, x = X, y = Y, 
                   balance_classes = T, model_id = "AD PRS full", 
                   nfolds = 5, fold_assignment = "Stratified", 
                   seed = seed_num, family = "binomial", lambda = 0, 
                   standardize = T, keep_cross_validation_models = F)
X = c("AD_PRS_rmapoe_norm", demo_set)
PRS.rmapoe = h2o.glm(training_frame = target.train, x = X, y = Y, 
                     balance_classes = T, model_id = "AD PRS without APOE", 
                     nfolds = 5, fold_assignment = "Stratified", 
                     seed = seed_num, family = "binomial", lambda = 0, 
                     standardize = T, keep_cross_validation_models = F)
X = c("e4count", demo_set)
e4count = h2o.glm(training_frame = target.train, x = X, y = Y, 
                  balance_classes = T, model_id = "e4count", 
                  nfolds = 5, fold_assignment = "Stratified", 
                  seed = seed_num, family = "binomial", lambda = 0, 
                  standardize = T, keep_cross_validation_models = F)
```

### 2) Direct epilepsy LASSO
```{r}
X = c(setdiff(names(genotype_df_map), mapped_AD_snp$uniqueID), demo_set)
snp.lasso.epi.feature = h2o.glm(training_frame = target.train, x = X, y = Y, 
                                balance_classes = T, model_id = "LASSO epi feature", 
                                nfolds = 5, fold_assignment = "Stratified", 
                                seed = seed_num, family = "binomial",
                                alpha = 1, lambda_search = TRUE,
                                standardize = T, keep_cross_validation_models = F)
snps_epilepsy = snp.lasso.epi.feature@model$variable_importances %>% 
  as.data.frame() %>% filter(variable %!in% demo_set) %>% 
  filter(scaled_importance > 0) %>% pull(variable)
X = c(snps_epilepsy, demo_set)
snp.direct.lasso.epi = h2o.glm(training_frame = target.train, x = X, y = Y, 
                               balance_classes = T, model_id = "LASSO direct epi", 
                               nfolds = 5, fold_assignment = "Stratified", 
                               seed = seed_num, family = "binomial", lambda = 0, 
                               standardize = T, keep_cross_validation_models = F)
```

### 3) Direct AD LASSO
```{r}
X = c(mapped_AD_snp$uniqueID, demo_set)
snp.lasso.epi.ADfeature = h2o.glm(training_frame = target.train, x = X, y = Y, 
                                balance_classes = T, model_id = "LASSO epi AD feature", 
                                nfolds = 5, fold_assignment = "Stratified", 
                                seed = seed_num, family = "binomial",
                                alpha = 1, lambda_search = TRUE,
                                standardize = T, keep_cross_validation_models = F)
snps_ADepilepsy = snp.lasso.epi.ADfeature@model$variable_importances %>% 
  as.data.frame() %>% filter(variable %!in% demo_set) %>% 
  filter(scaled_importance > 0) %>% pull(variable)
X = c(snps_ADepilepsy, demo_set)
snp.ADdirect.lasso.epi = h2o.glm(training_frame = target.train, x = X, y = Y, 
                                 balance_classes = T, model_id = "LASSO AD direct epi", 
                                 nfolds = 5, fold_assignment = "Stratified", 
                                 seed = seed_num, family = "binomial", lambda = 0, 
                                 standardize = T, keep_cross_validation_models = F)
```

### 4) Two-step LASSO
```{r}
X = c(mapped_AD_snp$uniqueID, demo_set, "pred_score")
snp.lasso2.epi.feature = h2o.glm(training_frame = target.train, x = X, y = Y, 
                                 balance_classes = T, model_id = "LASSO epi feature 2", 
                                 nfolds = 5, fold_assignment = "Stratified", 
                                 seed = seed_num, family = "binomial",
                                 alpha = 1, lambda_search = TRUE,
                                 standardize = T, keep_cross_validation_models = F)
snps_epi_specific = snp.lasso2.epi.feature@model$variable_importances %>% 
  as.data.frame() %>% filter(variable %!in% demo_set) %>% 
  filter(scaled_importance > 0) %>% pull(variable)
X = c(snps_epi_specific, demo_set)
snp.twostep.lasso.epi = h2o.glm(training_frame = target.train, x = X, y = Y, 
                                balance_classes = T, model_id = "LASSO two-step epilepsy", 
                                nfolds = 5, fold_assignment = "Stratified", 
                                seed = seed_num, family = "binomial", lambda = 0, 
                                standardize = T, keep_cross_validation_models = F)
```

```{r}
all_model_metric = c()
all_model_performance = c()
# Extract results
model_list = c("demo.only", "PRS.full", "PRS.rmapoe", "e4count",
               "snp.direct.lasso.epi", "snp.ADdirect.lasso.epi", "snp.twostep.lasso.epi")
for (i in 1:length(model_list)) {
  print(paste0("Model: ", model_list[i]))
  # Run extraction function
  extract_cv = extract_cv_results(get(model_list[i]))
  model_metric = t(extract_cv[[1]]) %>% as.data.frame() %>% 
    mutate(mark = model_list[i])
  model_performance = extract_cv[[2]] %>% as.data.frame() %>% 
    mutate(mark = model_list[i])
  if (i == 1) {
    all_model_metric = model_metric
    all_model_performance = model_performance
  } else {
    all_model_metric = rbind(all_model_metric, model_metric)
    all_model_performance = rbind(all_model_performance, model_performance)
  }
}
# Join results together
model_metric_epi = rbind(all_model_metric)
model_performance_epi = rbind(all_model_performance)
model_summary_epi = model_metric_epi %>% inner_join(model_performance_epi)
write.table(model_summary_epi,
            file = paste0(output_path, "model_summary_epi.txt"),
            sep = "\t", quote = F, row.names = F, col.names = T)
```

## 4. Save results
### Risk SNPs
```{r}
FUMA_data_path = paste0(raw_data_path, "FUMA/")
candidate_snps = read.table(file = paste0(FUMA_data_path, "FUMA_AD_new/snps.txt"), 
                            header = T, sep = "\t", fill = T)
load(file = paste0(raw_data_path, "modeling/EUR/mapped_snp.rda"))
candidate_snps_info = candidate_snps %>% 
  select(rsID, nearestGene, dist, func, posMapFilt, eqtlMapFilt, ciMapFilt)
rsID_check_AD = mapped_snp %>% 
  filter(uniqueID %in% c(snps_either, snps_AD_specific, snps_epi_specific)) %>% 
  select(rsID, uniqueID, chr, pos, a0, a1, gwasP, CADD, phenotype) %>% 
  filter(phenotype == "AD") %>% 
  mutate(snps_either = if_else(uniqueID %in% snps_either, 1, 0)) %>% 
  mutate(snps_AD_specific = if_else(uniqueID %in% snps_AD_specific, 1, 0)) %>% 
  mutate(snps_epi_specific = if_else(uniqueID %in% snps_epi_specific, 1, 0)) %>% 
  left_join(candidate_snps_info)
write.table(rsID_check_AD,
            file = paste0(output_path, "rsID_check_AD.txt"),
            sep = "\t", quote = F, row.names = F, col.names = T)

candidate_snps_epilepsy = read.table(file = paste0(FUMA_data_path, "FUMA_general_epilepsy_new/snps.txt"), 
                                     header = T, sep = "\t", fill = T)
candidate_snps_info_epi = candidate_snps_epilepsy %>% 
  select(rsID, nearestGene, dist, func, posMapFilt, eqtlMapFilt, ciMapFilt)
rsID_check_epi = mapped_snp %>% filter(uniqueID %in% snps_epilepsy) %>% 
  select(rsID, uniqueID, chr, pos, a0, a1, gwasP, CADD, phenotype) %>% 
  filter(phenotype == "epilepsy") %>% 
  mutate(snps_epilepsy = if_else(uniqueID %in% snps_epilepsy, 1, 0)) %>% 
  left_join(candidate_snps_info_epi)
write.table(rsID_check_epi,
            file = paste0(output_path, "rsID_check_epi.txt"),
            sep = "\t", quote = F, row.names = F, col.names = T)

save(rsID_check_AD, file = paste0(raw_data_path, "modeling/EUR/snps_interpret/rsID_check_AD.rda"))
save(rsID_check_epi, file = paste0(raw_data_path, "modeling/EUR/snps_interpret/rsID_check_epi.rda"))
# Output this for lead SNPs FUMA submission
candidate_snps_AD_short = candidate_snps %>% 
  select(rsID, chr, pos) %>% unique() 
lasso_either_snp = rsID_check_AD %>% filter(snps_either == 1) %>% select(rsID) %>% 
  left_join(candidate_snps_AD_short) %>% dplyr::rename("rsid" = "rsID") 
write.table(lasso_either_snp,
            file = paste0(output_path, "lasso_either_snp.txt"),
            sep = "\t", quote = F, row.names = F, col.names = T)
lasso_AD_snp = rsID_check_AD %>% filter(snps_AD_specific == 1) %>% select(rsID) %>% 
  left_join(candidate_snps_AD_short) %>% dplyr::rename("rsid" = "rsID") 
write.table(lasso_AD_snp,
            file = paste0(output_path, "lasso_AD_snp.txt"),
            sep = "\t", quote = F, row.names = F, col.names = T)
lasso_LOE_snp = rsID_check_AD %>% filter(snps_epi_specific == 1) %>% select(rsID) %>% 
  left_join(candidate_snps_AD_short) %>% dplyr::rename("rsid" = "rsID") 
write.table(lasso_LOE_snp,
            file = paste0(output_path, "lasso_LOE_snp.txt"),
            sep = "\t", quote = F, row.names = F, col.names = T)
```

### Predicted score dataframe
```{r}
step1_lasso_df = new_model_data %>% 
  rownames_to_column(var = "UniqueSampleId") %>% 
  mutate(UniqueSampleId = as.numeric(UniqueSampleId)) %>% 
  select(UniqueSampleId, pred_score, htn, diabetes, hyperlipid, either) %>% 
  left_join(patient_final_EUR)
# Cut with quintiles
step1_lasso_df$quantcut = quantcut(step1_lasso_df$pred_score, q = 5, na.rm = T)
explore_df_final = step1_lasso_df %>% 
  mutate(high_pred_score = if_else(quantcut == levels((step1_lasso_df$quantcut))[5], 1, 0),
         low_pred_score = if_else(quantcut == levels((step1_lasso_df$quantcut))[1], 1, 0)) %>% 
  mutate(condition = case_when(
    epilepsy == 0 & AD == 0 ~ "none",
    epilepsy == 1 & AD == 1 ~ "both",
    AD == 1 ~ "AD only",
    epilepsy == 1 ~ "LOE only",
  )) 
save(explore_df_final, file = paste0(raw_data_path, "modeling/EUR/explore_df_final.rda"))
```

## 5. SNP coefficients
```{r}
either_snp_coeff = SNP.lasso.either@model$coefficients_table
either_snp_coeff_top = either_snp_coeff %>% 
  filter(names %!in% demo_set) %>%
  arrange(desc(abs(standardized_coefficients))) %>% 
  filter(standardized_coefficients != 0) %>% 
  filter(names != "Intercept") %>% 
  left_join(rsID_check_AD, by = c("names" = "uniqueID")) %>% 
  mutate(coeff_cat = if_else(standardized_coefficients >= 0, "positive", "negative")) %>% 
  mutate(feature = case_when(
    names %in% demo_set ~ names,
    TRUE ~ rsID
  )) %>% filter(abs(standardized_coefficients) >= 0.01) %>% 
  mutate(group = case_when(
    !is.na(phenotype) ~ phenotype,
    names %in% demo_set ~ "demographics"
  )) %>% mutate(magnitude = abs(standardized_coefficients)) %>% 
  select(names, feature, standardized_coefficients, magnitude,
         phenotype, group, coeff_cat) 
# create a new factor variable for sorting the x-axis
either_snp_coeff_top$feature_sorted = reorder(either_snp_coeff_top$feature, 
                                              either_snp_coeff_top$magnitude)
# Plot the coefficients
pdf(paste0(output_path, "coef_plot.pdf"), width = 10, height = 5)
ggplot(either_snp_coeff_top, 
       aes(x = feature_sorted, y = magnitude, fill = coeff_cat)) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#4D4DF5", "#ED706B")) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 0.06)) + 
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
dev.off()
```
