---
title: "2_primary_anlaysis"
author: "Joy_Fu"
date: "2023-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls())
pacman::p_load(tidyverse, gtsummary, epiflow, gtools, caret, gridExtra, forestplot)
raw_data_path = "/Users/Mingzhou/Desktop/Projects/AD-epilepsy-genetics/data/"
output_path = "/Users/Mingzhou/Desktop/Projects/AD-epilepsy-genetics/output/"
# Source in useful functions
source("functions.R")

# Load in datasets
load(file = paste0(raw_data_path, "modeling/EUR/patient_final_EUR_04112023.rda"))
load(file = paste0(raw_data_path, "modeling/EUR/freeze60k_geno_freq.rda"))
load(file = paste0(raw_data_path, "modeling/EUR/genotype_df_map.rda"))
```

# Part 1. Descriptive statistics
```{r}
desc_prepare = patient_final_EUR %>% 
  mutate(htn = if_else(!is.na(HTNDate), 1, 0),
         diabetes = if_else(!is.na(DiabetesDate), 1, 0),
         stroke = if_else(!is.na(StrokeDate), 1, 0),
         hyperlipid = if_else(!is.na(HyperlipidDate), 1, 0),
         ad = if_else(!is.na(ADDate), 1, 0)) %>% 
  mutate(condition = case_when(
    epilepsy == 0 & ad == 0 ~ "none",
    epilepsy == 0 & ad == 1 ~ "AD only",
    epilepsy == 1 & ad == 0 ~ "epilepsy only",
    epilepsy == 1 & ad == 1 ~ "both"
  )) %>% 
  select(UniqueSampleId, ad, epilepsy, ad_age_diagnosis, epi_age_diagnosis, 
         age_last_visit, female, PC1, PC2, PC3, PC4, record_length, enc_per_yr, 
         htn, diabetes, stroke, hyperlipid, dementia, condition,
         e4count) 
# Desc by epilepsy status
desc_prepare %>% mutate(epi.cat = as.factor(epilepsy)) %>% 
  select(epi.cat, ad, epilepsy, ad_age_diagnosis, epi_age_diagnosis, 
         age_last_visit, female, record_length, enc_per_yr, 
         htn, diabetes, stroke, hyperlipid, dementia, condition, e4count) %>% 
  tbl_summary(by = epi.cat) %>% add_p()
```

```{r}
# Desc by ad status
desc_prepare %>% mutate(ad.cat = as.factor(ad)) %>% 
  select(ad.cat, ad, epilepsy, ad_age_diagnosis, epi_age_diagnosis, 
         age_last_visit, female, record_length, enc_per_yr, 
         htn, diabetes, stroke, hyperlipid, dementia, e4count) %>% 
  tbl_summary(by = ad.cat) %>% add_p()
```


# Part 2. Association tests
## 1. APOE (e4count)
```{r}
demo_basic = 'age_last_visit + female + PC1 + PC2 + PC3 + PC4 + record_length + enc_per_yr'
health_info = 'htn + diabetes + hyperlipid'

# Demographic only
demo_only = glm(epilepsy ~ age_last_visit + female + record_length + enc_per_yr, 
                data = desc_prepare, family = 'binomial')
# Adjusted for basic demo
e4count_demo = glm(as.formula(paste('epilepsy ~ e4count + ', demo_basic)), 
                   data = desc_prepare, family = 'binomial')
# Adjusted for health conditions
e4count_health = glm(as.formula(paste('epilepsy ~ e4count + ', demo_basic, ' + ', health_info)), 
                     data = desc_prepare, family = 'binomial')
```

## 2. AD PRS full


## 3. AD PRS w/o APOE


```{r message=FALSE, warning=FALSE}
# Run all models to get results
lst_model = c('demo_only', 'e4count_demo', 'e4count_health')
prs_cont_EUR = make_OR_table(lst_model, 2, 'prs_')
```



# Part 2. Important risk SNPs
## 1. Stepwise risky SNPs
```{r}
demo_set = c("female", "age_last_visit", "record_length", "enc_per_yr", 
             "PC1", "PC2", "PC3", "PC4")
stepwise_df_prep = freeze60k_geno_freq %>%
  mutate(ad = if_else(!is.na(ADDate), 1, 0)) %>%  
  select(all_of(names(genotype_df_map)), all_of(demo_set), epilepsy, ad, dementia)

f1 = paste(setdiff(names(stepwise_df_prep), c("epilepsy", "ad", "dementia")), 
           collapse = "` + `")
# Build forward selection model
intercept_only = glm(epilepsy ~ age_last_visit + female + PC1 + PC2 + PC3 + PC4 + 
                       record_length + enc_per_yr, 
                     data = stepwise_df_prep, family = 'binomial')
all = glm(paste0("epilepsy ~ `", f1, "`"), data = stepwise_df_prep, 
          family = 'binomial')
forward = step(intercept_only, direction = 'forward', 
               scope = formula(all), trace = 0)
selected_vars = gsub("\\+ ", "", forward$anova$Step)[2:length(forward$anova$Step)]
selected_vars_indsig = gsub("`", "", selected_vars)
print(selected_vars_indsig)
```

```{r}
FUMA_data_path = "/Users/Mingzhou/Desktop/Projects/Genetic-prediction-dementia/data/FUMA/FUMA_AD_EUR/"
candidate_snps = read.table(file = paste0(FUMA_data_path, "snps.txt"), 
                            header = T, sep = "\t", fill = T)
load(file = paste0(raw_data_path, "modeling/EUR/mapped_snp.rda"))
candidate_snps_info = candidate_snps %>% 
  select(rsID, nearestGene, dist, func, posMapFilt, eqtlMapFilt, ciMapFilt)
rsID_check = mapped_snp %>% filter(uniqueID %in% selected_vars_indsig) %>% 
  select(rsID, uniqueID, chr, pos, a0, a1, beta, se, gwasP, CADD) %>% 
  left_join(candidate_snps_info)

candidate_snps_short = candidate_snps %>% select(rsID, chr, pos)
mapped_stepwise_snp = rsID_check %>% select(rsID) %>% 
  left_join(candidate_snps_short) %>% dplyr::rename("rsid" = "rsID")
# write.table(mapped_stepwise_snp, 
#             file = paste0(output_path, "mapped_stepwise_snp.txt"),
#             sep = "\t", quote = F, row.names = F, col.names = T)
```

## 2. Get coefficients with 5-fold cross-validation
```{r}
# train-test split by group vars (epilepsy + AD)
stratified_data = stepwise_df_prep %>%
  rownames_to_column(var = "UniqueSampleId") %>% 
  group_by(epilepsy, ad) %>% group_split() 
set.seed(20230412)
none_group = stratified_data[[1]]
none_folds = createFolds(none_group$female, k = 5)
ad_group = stratified_data[[2]]
ad_folds = createFolds(ad_group$female, k = 5)
epilepsy_group = stratified_data[[3]]
epilepsy_folds = createFolds(epilepsy_group$female, k = 5)
both_group = stratified_data[[4]]
both_folds = createFolds(both_group$female, k = 5)

# Model preparation
demo_basic = 'age_last_visit + female + PC1 + PC2 + PC3 + PC4 + record_length + enc_per_yr'
f2 = paste(rsID_check$uniqueID, collapse = "` + `")
model_df_prep = freeze60k_geno_freq %>% select(all_of(rsID_check$uniqueID))

for (i in 1:5) {
  test_fold = rbind(none_group[none_folds[[i]], ],
                    ad_group[ad_folds[[i]], ],
                    epilepsy_group[epilepsy_folds[[i]], ],
                    both_group[both_folds[[i]], ])
  train_fold = stepwise_df_prep %>% 
    rownames_to_column(var = "UniqueSampleId") %>% 
    filter(UniqueSampleId %!in% test_fold$UniqueSampleId)
  
  # Fit model on training set only
  epilepsy_train = glm(as.formula(paste0('epilepsy ~ `', f2, '` + ', demo_basic)), 
                       data = train_fold, family = 'binomial')
  epilepsy_snp_coeff = epilepsy_train$coefficient %>% as.data.frame() %>% 
    rownames_to_column(var = "variable") %>% 
    filter(variable %!in% c(demo_set, "(Intercept)")) %>% 
    mutate(uniqueID = str_sub(variable, 2, -2)) %>% select(-variable)
  colnames(epilepsy_snp_coeff)[1] = "coefficient_epi"
  
  ad_train = glm(as.formula(paste0('ad ~ `', f2, '` + ', demo_basic)), 
                 data = train_fold, family = 'binomial')
  ad_snp_coeff = ad_train$coefficient %>% as.data.frame() %>% 
    rownames_to_column(var = "variable") %>% 
    filter(variable %!in% c(demo_set, "(Intercept)")) %>% 
    mutate(uniqueID = str_sub(variable, 2, -2)) %>% select(-variable)
  colnames(ad_snp_coeff)[1] = "coefficient_ad"
  # get the mean coefficient
  mean_coeff = epilepsy_snp_coeff %>% inner_join(ad_snp_coeff) %>% 
    mutate(coeff_mean = (coefficient_epi + coefficient_ad)/2)
  
  weighted_test = t(t(model_df_prep) * mean_coeff$coefficient_epi) %>% 
    as.data.frame() %>% mutate(shared_RS = rowSums(.)) %>% 
    rownames_to_column(var = "UniqueSampleId") %>% 
    mutate(UniqueSampleId = as.numeric(UniqueSampleId)) %>% 
    select(UniqueSampleId, shared_RS) %>% 
    filter(UniqueSampleId %in% test_fold$UniqueSampleId)
  
  if (i == 1) {
    weighted_test_combine = weighted_test
  } else {
    weighted_test_combine = rbind(weighted_test_combine, weighted_test)
  }
  
}
```

## 3. Association check (combined testing set)
```{r message=FALSE}
model_test_combine = weighted_test_combine %>% 
  inner_join(desc_prepare) %>% inner_join(patient_final_EUR) 

df_for_cluster = model_df_prep %>% rownames_to_column(var = "UniqueSampleId") %>% 
  mutate(UniqueSampleId = as.numeric(UniqueSampleId)) %>% 
  inner_join(desc_prepare) %>% inner_join(patient_final_EUR) %>% 
  mutate(condition = case_when(
    epilepsy == 0 & ad == 0 ~ "none",
    epilepsy == 0 & ad == 1 ~ "AD only",
    epilepsy == 1 & ad == 0 ~ "epilepsy only",
    epilepsy == 1 & ad == 1 ~ "both"
  ))
```

```{r}
# Fit models in testing sets
demo_basic = 'age_last_visit + female + PC1 + PC2 + PC3 + PC4 + record_length + enc_per_yr'
health_info = 'htn + diabetes + hyperlipid'

# Association with epilepsy
epilepsy_demo = glm(as.formula(paste('epilepsy ~ shared_RS + ', demo_basic)), 
                    data = model_test_combine, family = 'binomial')
epilepsy_health = glm(as.formula(paste('epilepsy ~ shared_RS + ', demo_basic, 
                                       ' + ', health_info)), 
                    data = model_test_combine, family = 'binomial')
# Association with AD
ad_demo = glm(as.formula(paste('ad ~ shared_RS + ', demo_basic)), 
              data = model_test_combine, family = 'binomial')
ad_health = glm(as.formula(paste('ad ~ shared_RS + ', demo_basic, ' + ', health_info)), 
                data = model_test_combine, family = 'binomial')
# Association with dementia
dem_demo = glm(as.formula(paste('dementia ~ shared_RS + ', demo_basic)), 
               data = model_test_combine, family = 'binomial')
dem_health = glm(as.formula(paste('dementia ~ shared_RS + ', demo_basic, ' + ', health_info)), 
                 data = model_test_combine, family = 'binomial')
```

```{r message=FALSE, warning=FALSE}
# Run all models to get results
lst_model = c('epilepsy_demo', 'epilepsy_health', 'ad_demo', 'ad_health',
              'dem_demo', 'dem_health')
sharedRS_test = make_OR_table(lst_model, 2, 'sharedRS_')
```

## 4. Final train on the whole set
```{r}
# Fit model on training set only
epilepsy_full = glm(as.formula(paste0('epilepsy ~ `', f2, '` + ', demo_basic)), 
                    data = stepwise_df_prep, family = 'binomial')
epilepsy_snp_coeff_final = coef(summary(epilepsy_full)) %>% as.data.frame() %>% 
  rownames_to_column(var = "variable") %>% 
  filter(variable %!in% c(demo_set, "(Intercept)")) %>% 
  mutate(uniqueID = str_sub(variable, 2, -2)) %>% select(uniqueID, Estimate, `Std. Error`)
colnames(epilepsy_snp_coeff_final) = c("uniqueID", "coefficient", "std")
  
ad_full = glm(as.formula(paste0('ad ~ `', f2, '` + ', demo_basic)), 
              data = stepwise_df_prep, family = 'binomial')
ad_snp_coeff_final = coef(summary(ad_full)) %>% as.data.frame() %>% 
  rownames_to_column(var = "variable") %>% 
  filter(variable %!in% c(demo_set, "(Intercept)")) %>% 
  mutate(uniqueID = str_sub(variable, 2, -2)) %>% select(uniqueID, Estimate, `Std. Error`)
colnames(ad_snp_coeff_final) = c("uniqueID", "coefficient", "std")
# get the mean coefficient
mean_coeff_final = epilepsy_snp_coeff_final %>% inner_join(ad_snp_coeff_final) %>% 
  mutate(coeff_mean = (coefficient_epi + coefficient_ad)/2) %>% 
  left_join(rsID_check) %>% 
  select(rsID, chr, pos, a0, a1, beta, se, gwasP, CADD, 
         posMapFilt, eqtlMapFilt, ciMapFilt,
         coefficient_ad, coefficient_epi, coeff_mean) %>% 
  mutate(mean_cat = if_else(coeff_mean >= 0, "positive", "negative"),
         ad_cat = if_else(coefficient_ad >= 0, "positive", "negative"),
         epi_cat = if_else(coefficient_epi >= 0, "positive", "negative"))
  
weighted_full = t(t(model_df_prep) * coeff_lst) %>% 
  as.data.frame() %>% mutate(shared_RS = rowSums(.)) %>% 
  rownames_to_column(var = "UniqueSampleId") %>% 
  mutate(UniqueSampleId = as.numeric(UniqueSampleId)) %>% 
  select(UniqueSampleId, shared_RS) 

explore_df_final = weighted_full %>% 
  inner_join(desc_prepare) %>% inner_join(patient_final_EUR) 
dim(explore_df_final) # dim = (18742,53)
```

```{r}
# Plot the coefficients
plot_mean = ggplot(mean_coeff_final, aes(x = rsID, y = coeff_mean, fill = mean_cat)) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#4D4DF5", "#ED706B")) +
  coord_flip() +
  scale_y_continuous(limits = c(-0.4, 0.45)) + 
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
plot_epi = ggplot(mean_coeff_final, aes(x = rsID, y = coefficient_epi, fill = epi_cat)) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#4D4DF5", "#ED706B")) +
  coord_flip() +
  scale_y_continuous(limits = c(-0.4, 0.45)) + 
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
plot_ad = ggplot(mean_coeff_final, aes(x = rsID, y = coefficient_ad, fill = ad_cat)) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#4D4DF5", "#ED706B")) +
  coord_flip() + 
  scale_y_continuous(limits = c(-0.4, 0.45)) +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
```

```{r}
pdf(paste0(output_path, "coef_plot.pdf"), width = 15, height = 5)
grid.arrange(plot_epi, plot_ad, plot_mean, nrow = 1)
dev.off()
```


# Part 3. Explore shared risk score
```{r}
levels(quantcut(explore_df_final$shared_RS, q = 5, na.rm = T))
explore_df_final = explore_df_final %>% 
  mutate(high_sharedRS = if_else(shared_RS > 0.559, 1, 0),
         low_sharedRS = if_else(shared_RS <= 0.053, 1, 0)) %>% 
  mutate(condition = case_when(
    epilepsy == 0 & ad == 0 ~ "none",
    epilepsy == 0 & ad == 1 ~ "AD only",
    epilepsy == 1 & ad == 0 ~ "epilepsy only",
    epilepsy == 1 & ad == 1 ~ "both"
  ))
explore_df_short = explore_df_final %>% 
  filter(high_sharedRS == 1 | low_sharedRS == 1)
```

## Top 20% vs. bottom 20% score
```{r}
explore_df_short %>% mutate(high_sharedRS.cat = as.factor(high_sharedRS),
                      female.cat = as.factor(female)) %>%
  mutate(n_conditions = epilepsy + dementia) %>% 
  dplyr::select(high_sharedRS.cat, age_last_visit, female.cat, record_length, enc_per_yr, 
         epilepsy, e4count, dementia, htn, diabetes, stroke, hyperlipid, ad, 
         epi_age_diagnosis, dem_age_diagnosis, ad_age_diagnosis, n_conditions,
         condition) %>% 
  tbl_summary(by = high_sharedRS.cat) %>% add_p()
```

## Number of conditions
```{r}
explore_df_final %>% 
  mutate(high_sharedRS.cat = as.factor(high_sharedRS),
         female.cat = as.factor(female)) %>%
  dplyr::select(high_sharedRS.cat, age_last_visit, female.cat, record_length, 
         enc_per_yr, epilepsy, e4count, dementia, htn, diabetes, stroke, 
         hyperlipid, ad, epi_age_diagnosis, dem_age_diagnosis, ad_age_diagnosis, 
         condition, shared_RS) %>% 
  tbl_summary(by = condition) %>% add_p()
```

## Forest plot
```{r}
var_names = c('epilepsy', 'ad', 'dementia', # 3
              'age_last_visit', 'female', 'record_length', 'enc_per_yr', # 4
              'htn', 'diabetes', 'stroke', 'hyperlipid', # 4
              'e4count') # 1
explanary_names = c('Late-onset Epilepsy', "Alzheimer's Disease", 'Dementia', 
                    'Age at last visit', 'Sex (female)', 'EHR record length', 
                    'EHR record density', 'Hypertension', 'Diabetes', 'Stroke', 
                    'Hyperlipid', 'APOE-e4 allele count')

model_lst = list()
for (i in 1:length(var_names)) {
  model_lst[[i]] = lm(as.formula(paste0('shared_RS ~ ', var_names[i], 
                                        " + PC1 + PC2 + PC3 + PC4")),
                      data = explore_df_short)
}
table1 = build_coef_table(var_names, model_lst)

forest.data = table1 %>% 
  dplyr::select(beta, lower_CI, upper_CI) %>% 
  mutate(sub = c('Primary outcomes', rep(NA, 2), 
                 'Demographics', rep(NA, 3), 
                 'Health conditions' , rep(NA, 3),
                 'AD genetics', rep(NA, 0))) %>% 
  mutate(beta = round(as.numeric(as.character(beta)), 3),
         lower_CI = round(as.numeric(as.character(lower_CI)), 3),
         upper_CI = round(as.numeric(as.character(upper_CI)), 3)) %>% 
  mutate(class = c(rep(2, 3), rep(3, 4), rep(4, 4), rep(5, 1)))

tabletext = cbind(
  c("Category", "\n", forest.data$sub),
  c("Exposure", "\n", explanary_names),
  c("Beta coefficient (95% CI)", "\n", as.character(table1$text))
)
pop.cols = c("black","black","black", "black", "black")
```

```{r}
pdf(paste0(output_path, "Shared_RS_Forest.pdf"), width = 10, height = 5)
forestplot(labeltext = tabletext, graph.pos = 1,
           mean = c(NA, NA, forest.data$beta), 
           lower = c(NA, NA, forest.data$lower_CI),
           upper = c(NA, NA, forest.data$upper_CI),
           xticks = c(-0.1, 0, 0.1, 0.2, 0.3, 0.4),
           zero = 0,
           title = "Figure 4. Associations between covariates and shared genetic risk score (n = 18,742)",
           xlab = "Effect Size",
           txt_gp = fpTxtGp(label = list(gpar(fontface = "bold", cex = 0.8, 
                                              fontfamily = "serif"),
                                         gpar(cex = 0.8, fontfamily = "serif"),
                                         gpar(cex = 0.8, fontfamily = "serif")),
                            ticks = gpar(cex = 0.6, fontfamily = "serif"),
                            xlab = gpar(cex = 0.7, fontfamily = "serif"),
                            title = gpar(cex = 1, fontfamily = "serif")),
           col = fpColors(text = pop.cols[c(1, 1, forest.data$class)], 
                          box  ="black",
                          lines = "black", 
                          zero ="gray50"),
           cex = 0.2, lineheight = "auto", boxsize = 0.25, 
           lwd.ci = 1, ci.vertices = TRUE, ci.vertices.height = 0.15)
dev.off()
```


## Causal inference
### 1) New dataset (onset before)
```{r}
causation_prep = explore_df_final %>% 
  mutate(epilepsy_before = case_when(
    epilepsy == 1 & !is.na(ad_age_diagnosis) & (epi_age_diagnosis < ad_age_diagnosis) ~ 1,
    epilepsy == 1 & is.na(ad_age_diagnosis) ~ 1,
    TRUE ~ 0
  )) %>% 
  mutate(ad_before = case_when(
    ad == 1 & !is.na(epi_age_diagnosis) & (ad_age_diagnosis < epi_age_diagnosis) ~ 1,
    ad == 1 & is.na(epi_age_diagnosis) ~ 1,
    TRUE ~ 0
  )) %>% 
  mutate(diff_year = ad_age_diagnosis - epi_age_diagnosis)
table(causation_prep$epilepsy, causation_prep$epilepsy_before, useNA = "ifany")
table(causation_prep$ad, causation_prep$ad_before, useNA = "ifany")

```



```{r}
ad_epi = glm(paste0('ad ~ epilepsy_before + ', demo_basic, 
                      ' + ', health_info), 
              data = causation_prep, family = 'binomial')
summary(ad_epi)
```

```{r}
ad_epi_gen = glm(paste0('ad ~ epilepsy_before + shared_RS + ', demo_basic, 
                      ' + ', health_info), 
              data = causation_prep, family = 'binomial')
summary(ad_epi_gen)
```


```{r}
epi_ad = glm(paste0('epilepsy ~ ad_before + shared_RS +', demo_basic, 
                      ' + ', health_info), 
              data = causation_prep, family = 'binomial')
summary(epi_ad)
```


```{r}
names(explore_df_final)
```

